---
title: "Highlights of the Syllabuses (Syllabi)?"
subtitle: "EFB 796/797: Concepts & Techniques in Movement and Spatial Ecology"
author: "Elie Gurarie and Nate Wehr"
date: "August 25, 2025"
output: 
  xaringan::moon_reader:
    css: ["default", "default-fonts", "mycss.css"]
    nature:
      highlightStyle: github
      countIncrementalSlides: false
      highlightLines: true
      ratio: '16:9'
editor_options: 
  chunk_output_type: console
---

<!-- https://bookdown.org/yihui/rmarkdown/xaringan-format.html -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, 
                      warning = FALSE, las = 1)
#output: html_document
```


```{r xaringan-scribble, echo=FALSE}
xaringanExtra::use_scribble()
```

```{r xaringan-tile-view, echo=FALSE}
xaringanExtra::use_tile_view()
```

```{r colsFunction, eval = FALSE, echo = FALSE}
#system("cp ../mycss.css ./")
xaringan::inf_mr()
```



.pull-left[

## 797: Concepts 
### 1 credit seminar

- in-depth discussions of concepts important to movement (& spatial) ecology
  - 1-2 papers a week, with input on selection from students. 

- reading materials relevant to graduate work. 

Collegial discussion format to improve understanding of the underlying premises and applications of these methodological approaches.

]

--

.pull-right[

## 796: Techniques
### 2 credit lecture / lab

Linking ecological **questions** to **statistical tools**

- Learning to deal with data

- Understanding underlying math / stats.

- Visualizing, summarizing, inference

- Moving **your own projects** forward. 

]
--

<br> 
### Combined:

- Papers covering methods and applications of interest to **796 students** will be covered in **797**.  


---


#  Assessment

.pull-left.large[
## 797 only

- Lead a discussion 

- Writing assignment 
]

.pull-right.large[
## 796 (+797)

- Lead a discussion

- Some homework assignments / exercises

- Final analysis project presentation 

]
